Forcing for Isabelle

### 2019-01-05 LICS/proof-of-separation.tex
eds. varias. esp. recursion
comiendo de powerset en paper

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Proof of Separation}
\label{sec:proof-separation}

This proof can be found in the file \verb|Separation_Axiom.thy| of the
development, which we proceed to discuss in detail. The order chosen to
implement
the proof sought to minimize the cross-reference of facts;  it is
not entirely appropriate for a text version, so we depart from it in
this presentation. Nevertheless, we will refer to each specific block
of code by line number for ease of reference.

@@

Since $m$ is a name defined using Separation, we may use
(\ref{eq:val-name-sep}) to show (lines 221--274 of
\isatt{Separation\_Axiom})
\[
  \val(G,m) = \{x\in c : M[G],[x,w,c]\models \phi(x_0,x_1)\}.
\]
The right-hand side is trivially equal to $S$, but as a consequence of
the definition of 
\isatt{separation\_ax}, the result contains an extra $c$ in the
environment.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%          
\section{Separation and Powerset}
\label{sec:proof-separation}
We proceed to describe in detail the main goal of this paper, the formalization
of the proof of the Separation Axiom. Afterwards, we sketch  the
implementation of the Powerset Axiom.

This proof of Separation can be found in the file
\verb|Separation_Axiom.thy|. The order chosen to
implement
the proof sought to minimize the cross-reference of facts;  it is
not entirely appropriate for a text version, so we depart from it in
this presentation. Nevertheless, we will refer to each specific block
of code by line number for ease of reference.

@@

Since $m$ is a name defined using Separation, we may use
(\ref{eq:val-name-sep}) to show (lines 221--274 of
\isatt{Separation\_Axiom})
\begin{equation}
  \label{eq:val-of-m}
  \val(G,m) = \{x\in c : M[G],[x,w,c]\models \phi(x_0,x_1)\}.
\end{equation}
The right-hand side is trivially equal to $S$, but as a consequence of
the definition of 
\isatt{separation\_ax}, the result contains an extra $c$ in the
environment.

@@

We now turn to the Powerset Axiom. We followed the proof of
\cite[IV.2.27]{kunen2011set}, and actually the main technical result,
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ Pow{\isacharunderscore}inter{\isacharunderscore}MG{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}a{\isasymin}M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}Pow{\isacharparenleft}a{\isacharparenright}\ {\isasyminter}\ M{\isacharbrackleft}G{\isacharbrackright}\ {\isasymin}\ M{\isacharbrackleft}G{\isacharbrackright}{\isachardoublequoteclose}
\end{isabelle}
keeps most of the structure of the printed proof; this ``skeleton'' of the
argument takes around 120 (short) lines. There are approximately 30
more lines of bureaucracy in the proof of the last lemma. We preserved the
names of variables used in the textbook, with the occasional question
mark that distinguishes meta-variables.

Two more
absoluteness results concerning powersets were needed: These are
refinements of results located in \isatt{Relative} that assume that
$y$ below is in the class $M$.
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ {\isacharparenleft}\isakeyword{in}\ M{\isacharunderscore}trivial{\isacharparenright}\ powerset{\isacharunderscore}subset{\isacharunderscore}Pow{\isacharcolon}\isanewline
\ \ \isakeyword{assumes}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}powerset{\isacharparenleft}M{\isacharcomma}x{\isacharcomma}y{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymAnd}z{\isachardot}\ z{\isasymin}y\ {\isasymLongrightarrow}\ M{\isacharparenleft}z{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\ \isanewline
\ \ \ \ {\isachardoublequoteopen}y\ {\isasymsubseteq}\ Pow{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}
\end{isabelle}
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ {\isacharparenleft}\isakeyword{in}\ M{\isacharunderscore}trivial{\isacharparenright}\ powerset{\isacharunderscore}abs{\isacharcolon}\ \isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}M{\isacharparenleft}x{\isacharparenright}{\isachardoublequoteclose}\ {\isachardoublequoteopen}{\isasymAnd}z{\isachardot}\ z{\isasymin}y\ {\isasymLongrightarrow}\ M{\isacharparenleft}z{\isacharparenright}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ {\isachardoublequoteopen}powerset{\isacharparenleft}M{\isacharcomma}x{\isacharcomma}y{\isacharparenright}\ {\isasymlongleftrightarrow}\ y\ {\isacharequal}\ {\isacharbraceleft}a{\isasymin}Pow{\isacharparenleft}x{\isacharparenright}\ {\isachardot}\ M{\isacharparenleft}a{\isacharparenright}{\isacharbraceright}{\isachardoublequoteclose}
\end{isabelle}
